---
title: "Modeling and prediction for movies"
output:
  pdf_document: default
  html_document:
    fig_height: 4
    highlight: pygments
    theme: spacelab
---

## Setup

### Load packages

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
library(statsr)
library(corrplot)
```

### Load data

Make sure your data and R Markdown files are in the same directory. When loaded
your data file will be called `movies`. Delete this note when before you submit 
your work. 

```{r load-data}
load("movies.Rdata")
```



* * *

## Part 1: Data


The data set consists of 651 movies produced and released before 2016. The source of this data comes from Rotten Tomatoes and IMDb (Internet Movie Database). According to the information given, the data set is randomly sampled, meaning the sample can be generalized to the population. However, there is no mention of random assignment, so causation cannot be established as a part of this study.


* * *

## Part 2: Research question


For this project, I will determine which attributes, from the dataset provided, make a movie popular.


Additionally, through exploratory data analysis, I will determine if there is a possible relationship between the movie's popularity, and whether it won an Oscar award or not.


* * *

## Part 3: Exploratory data analysis


First, I shall look into the relationship between a movie's popularity and if the movie had won an oscar.
To determine what makes a movie popular, I have created a variable *score* which is the average of the rating on IMDB and the audience score on Rotten Tomatoes. I have not considered critic score as I am considering only audience ratings as a measure of popularity.


### Creating the score variable


Lets take a look at the `audience_score` and `imdb_rating`:


```{r}
movies %>% select(imdb_rating, audience_score)
```


```{r}
summary(movies$audience_score)
summary(movies$imdb_rating)
```


For these two variables to be combined they need to be on the same scale. This can be achieved by multiplying the values in the `imdb_rating` column by 10.


I will create a variable `score` which will be the average of the `imdb_rating` and `audience_score`.


```{r}
movie_score = ((movies$imdb_rating*10)+movies$audience_score)/2
summary(movie_score)
```


### Plot of Oscar nominees and Oscar winners


Taking a subset from the dataset, I will show the distribution of movies that have won an Oscar.


```{r}
movies_1 <- movies %>% mutate(score = movie_score)
oscar_score <- movies_1 %>% select(audience_score, imdb_rating, best_pic_nom, best_pic_win, score)

oscar_score
```


```{r}
oscar_score %>% 
  group_by(best_pic_nom, best_pic_win) %>% 
  summarise(count = n())
```


There seems to be a data issue as one of the records say that it was not nominated for an Oscar but it won the award. This is most likely an issue in the dataset.


```{r}
movies_1 %>% 
  filter(best_pic_nom == 'no', best_pic_win == 'yes') %>% 
  select(title)
```


According to IMDB, The Hurt Locker has won an Oscar for best picture, so I can rectify this issue in the data.


```{r}
movies_1 <- movies_1 %>% 
  mutate(best_pic_nom = replace(best_pic_nom, best_pic_nom=="no" & best_pic_win=="yes","yes"))
```


Now taking a subset of the data again, we can see that the data inaccuracy has been handled.


```{r}
oscar_score <- movies_1 %>% 
  select(audience_score, imdb_rating, best_pic_nom, best_pic_win) %>% 
  mutate(score = movie_score)

oscar_score %>% 
  group_by(best_pic_nom, best_pic_win) %>% 
  summarise(count = n())
```


Now onto visualizating the Oscar winners and their scores.


```{r}
ggplot(oscar_score, aes(x = best_pic_win, y = score)) + 
  geom_point(aes(color=best_pic_nom)) + theme_minimal()
```


We can see that movies that were given a high score won the Oscar award. To see if these two groups are significantly different, we can visualize with a boxplot.


```{r}
ggplot(oscar_score, aes(x = best_pic_win, y = score)) + 
  geom_boxplot() + theme_minimal()
```


Since the two boxes are not intersecting with eachother, we can say that the two groups **are** different from eachother.


### Cleaning data


Lets check which columns in the dataset have missing values.


```{r}
colnames(movies_1)[colSums(is.na(movies_1)) > 0]
```


The columns `studio`, `director`, and actor columns from 1 to 5 do not seem to be relevant to our task so I will not be considering them in the data cleaning process.


```{r}
summary(movies_1[c('runtime','dvd_rel_year','dvd_rel_month','dvd_rel_day')])
```


Lets take a look at these NA rows in more detail.


```{r}
movies_1 %>% 
  filter(is.na(dvd_rel_year) | is.na(dvd_rel_month) | is.na(dvd_rel_day) | is.na(runtime)) %>% 
  select(title, runtime, dvd_rel_year, dvd_rel_month, dvd_rel_day)
```


Initially, I had planned to impute these missing values based on information from IMDB. However, the movies with NA values in columns related to the DVD release date were released before DVD was introduced.


```{r}
movies_1 %>% 
  filter(is.na(dvd_rel_year)) %>% 
  select(title, thtr_rel_year, thtr_rel_month, thtr_rel_day)
```


**Charlie: The Life and Art of Charles Chaplin** is the only row where the DVD release date is available. So I have decided to impute values for this movie, and **The End of America**. The rest of the rows will be dropped.


```{r}
movies_2 <- movies_1 %>% 
  mutate(runtime = replace(runtime, title=="The End of America",74),
      dvd_rel_year = replace(dvd_rel_year, title == "Charlie: The Life and Art of Charles Chaplin", 2003),
      dvd_rel_month = replace(dvd_rel_month, title == "Charlie: The Life and Art of Charles Chaplin", 11),
      dvd_rel_day = replace(dvd_rel_day, title == "Charlie: The Life and Art of Charles Chaplin", 3))

movies_2 %>% 
  filter(title == "The End of America" | title == "Charlie: The Life and Art of Charles Chaplin") %>% 
  select(title, runtime, dvd_rel_year, dvd_rel_month, dvd_rel_day)
```


Now I shall drop the remaining rows with NA.


```{r}
movies_3 <- movies_2 %>% filter(!is.na(dvd_rel_year), !is.na(dvd_rel_month), !is.na(dvd_rel_day))

head(movies_3)
```


### Scatter plots


```{r}
ggplot(movies_3, aes(x = runtime, y = score)) + 
  geom_point() + stat_smooth(method=lm) + theme_minimal()
```


Looking at the plot above, there seems to be a very weak relationship between `runtime` and `score`.


```{r}
ggplot(movies_3, aes(x = critics_score, y = score)) + 
  geom_point() + stat_smooth(method=lm) + theme_minimal()
```


For the plot of `critics_score` and `score`, there seems to be a moderately strong relationship between the two variables.


```{r}
ggplot(movies_3, aes(x = imdb_num_votes, y = score)) + 
  geom_point() + stat_smooth(method=lm) + theme_minimal()
```


There seems to be an extreme right skewness in the imdb_num_votes variable. A weak linear relationship may exist between the two variables.


### Correlation Matrix


For the correlation matrix, I shall extract the numerical columns from the dataset. Note that I am leaving out `imdb_rating` and `audience_score` as these both were used to make the `score` variable. It is expected that the `score` variable would high very high correlation with the two other variables, and this might be undesirable for our model.


```{r}
nums <- unlist(lapply(movies_3, is.numeric))
numerical_cols <- movies_3[,nums]
numerical_cols <- select(numerical_cols, -c(8,11))
```


Now plotting the correlation matrix:


```{r}
cor_mat <- cor(numerical_cols)

corrplot(cor_mat, type="upper", method="number")
```


There only seems to be a strong correlation with `critics_score`. The two other variables we plotted above, `imdb_num_votes` and `runtime` have a very weak positive correlation with score.
Additionally, there seems to be a moderately strong positive relatonship between `thtr_rel_year` and `dvd_rel_year`. Even though this relationship is not strong, it is in our best interest to remove this variable to prevent any collinearity in the model.


* * *

## Part 4: Modeling


For modeling, I will be considering the following variables:

`title_type`, `genre`, `runtime`, `mpaa_rating`, `thtr_rel_year`, `thtr_rel_month`, `thtr_rel_day`, `dvd_rel_month`, `dvd_rel_day`, `imdb_num_votes`, `critics_rating`, `critics_score`, `audience_rating`, `best_pic_nom`, `best_pic_win`, and `top200_box`.

In addition to variables related to 'score' being removed, variables such as the name of the movie, directors, or actors have been removed as well since they don't seem to be meaningful in our model. URLs to the movie page has been removed as well.

Forward selection will be implemented in this model to make sure our model has only the relevant variables, making it parsimonious. We will refit the model, removing each variable until we get the highest R squared value. For convenience of the readers, the working has been omitted from this markdown file.


### Model Selection


```{r}
model_1 <- lm(score ~ audience_rating, data=movies_3)

summary(model_1)$adj.r.squared
```


```{r}
model_2 <- lm(score ~ audience_rating + critics_score, data=movies_3)

summary(model_2)$adj.r.squared
```


```{r}
model_3 <- lm(score ~ audience_rating + critics_score + imdb_num_votes, data=movies_3)

summary(model_3)$adj.r.squared
```


```{r}
model_4 <- lm(score ~ audience_rating + critics_score + imdb_num_votes + genre, data=movies_3)

summary(model_4)$adj.r.squared
```


```{r}
model_5 <- lm(score ~ audience_rating + critics_score + imdb_num_votes + genre + critics_rating, data=movies_3)

summary(model_5)$adj.r.squared
```


```{r}
model_6 <- lm(score ~ audience_rating + critics_score + imdb_num_votes + genre + critics_rating + thtr_rel_year, data=movies_3)

summary(model_6)$adj.r.squared
```


```{r}
model_7 <- lm(score ~ audience_rating + critics_score + imdb_num_votes + genre + critics_rating + thtr_rel_year + dvd_rel_month, data=movies_3)

summary(model_7)$adj.r.squared
```


```{r}
model_8 <- lm(score ~ audience_rating + critics_score + imdb_num_votes + genre + critics_rating + thtr_rel_year + dvd_rel_month + best_pic_nom, data=movies_3)

summary(model_8)$adj.r.squared
```


```{r}
model_9 <- lm(score ~ audience_rating + critics_score + imdb_num_votes + genre + critics_rating + thtr_rel_year + dvd_rel_month + best_pic_nom + title_type, data=movies_3)

summary(model_9)$adj.r.squared
```


```{r}
model_10 <- lm(score ~ audience_rating + critics_score + imdb_num_votes + genre + critics_rating + thtr_rel_year + dvd_rel_month + best_pic_nom + title_type + top200_box, data=movies_3)

summary(model_10)$adj.r.squared
```


```{r}
model_11 <- lm(score ~ audience_rating + critics_score + imdb_num_votes + genre + critics_rating + thtr_rel_year + dvd_rel_month + best_pic_nom + title_type + top200_box + runtime, data=movies_3)

summary(model_11)$adj.r.squared
```


```{r}
model_12 <- lm(score ~ audience_rating + critics_score + imdb_num_votes + genre + critics_rating + thtr_rel_year + dvd_rel_month + best_pic_nom + title_type + top200_box + runtime + best_pic_win, data=movies_3)

summary(model_12)$adj.r.squared
```


This is the final model as addition of any more variables will not increase the adjust R squared value


```{r}
final_model <- model_12

summary(final_model)
```


### Model Diagnostics


#### Linear Relationship between Explanatory and Response variable


```{r}
plot(final_model$residuals ~ movies_3$critics_score) +abline(h = 0)
plot(final_model$residuals ~ movies_3$imdb_num_votes) +abline(h = 0)
plot(final_model$residuals ~ movies_3$thtr_rel_year) +abline(h = 0)
plot(final_model$residuals ~ movies_3$dvd_rel_month) +abline(h = 0)
plot(final_model$residuals ~ movies_3$runtime) +abline(h = 0)
```


There only seems to be an issue in the `imdb_num_votes` plot. All other plots seem to meet the condition.


```{r}
hist(final_model$residuals, prob=TRUE)
lines(density(final_model$residuals), col="blue", lwd = 2)
qqnorm(final_model$residuals)
qqline(final_model$residuals)
```


Majority of the points are on the line except for points at the tails, this indicates some skewness. Additionally, the histogram appears to be nearly normal.


```{r}
plot(final_model$residuals ~ final_model$fitted) + abline(h=0, col="blue")
plot(abs(final_model$residuals) ~ final_model$fitted) + abline(h=0, col="blue")
```


Plot of fitted values shows the residuals are equally variable.


#### Coefficients Interpretation


Some of the coefficients interpretations are given below.

`audience_ratingUpright`: All else held constant, the model predicts that movies rated 'Upright' will have a score 16.91 higher on average than movies rated 'Spilled'.

`critics_score`: All else held constant, for every unit increase in `critics_score`, the model predicts a 0.237 decrease in score on average.

`genreAnimation`: All else held constant, the model predicts that movies of genre 'Animation' will have a score 1.17 lower on average than movies of genre 'Action & Animation'.

`critics_ratingFresh`: All else held constant, the model predicts that movies given a critics rating of 'Fresh' will have a score 1.268 higher on average than movies given a critics rating of 'Certified Fresh'.

`title-typeTV Movie`: All else held constant, the model predicts that movies of type 'TV Movie' will have a score 3.85 lower on average than movies of type 'Documentary'.


* * *

## Part 5: Prediction


I will be predicting the hit film "Captain America: Civil War" with this model.
References: [IMDB](https://www.imdb.com/title/tt3498820/) [Rotten Tomatoes](https://www.rottentomatoes.com/m/captain_america_civil_war)


I have calculated the score below, using the data from both pages.


```{r}
cap_score = ((7.8*10)+89)/2
cap_score
```


Creating test data for the model.


```{r}
audience_rating <- "Upright"
critics_score <- 91
imdb_num_votes <- 639952
genre <- "Action & Adventure"
critics_rating <- "Certified Fresh"
thtr_rel_year <- 2016
dvd_rel_month <- 9
best_pic_nom <- "no"
title_type <- "Feature Film"
top200_box <- "yes"
runtime <- 147
best_pic_win <- "no"

test <- data.frame(audience_rating, critics_score, imdb_num_votes, genre,
                   critics_rating, thtr_rel_year, dvd_rel_month, best_pic_nom,
                   title_type, top200_box, runtime, best_pic_win)
```


```{r}
predict(final_model, test, interval = "prediction", level = 0.95)
```


We are 95% confident that, all else held equal, the predicted score for "Captain America: Civil War" will be between 71.097 and 97.583 on average.


* * *

## Part 6: Conclusion


I have created a model using forward selection and adjusted R Squared as our selection criteria. Although we managed to get an accurate prediction, the margin of error is large (97.58 - 71.1 = 26.48). This could be due to the way the `score` variable was created. Additionally, it would be best to refit the model without any time related variables. Additionally, we may get valuable insights on the data if our dataset had a distinction for actors who won an Oscar prior to being cast in the movie, and actors who won an Oscar due to their role in the movie.

